---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Water Discharge Analysis R Package.

This package was developed to help analyze stream discharge data. 
It is an alternative to using the HEC package.

This package was written for the B.C. Ministry of Environment
by Carl James Schwarz (cschwarz.stat.sfu.ca@gmail.com).


### Installing the package.

To install the latest version of BCWaterDischargeAnalysis:


```{r, eval = FALSE}
library(devtools)   # has the install_github() function
# This will take about 5 minutes to install
install_github("bcgov/water-discharge-analysis/BCWaterDischargeAnalysis", build_vignettes = TRUE) 
# Using `build_vignettes = TRUE` will slow down the install, but is necessary if 
# you want to read the vignette, which is recommended
library(BCWaterDischargeAnalysis)
```

You can see the vignette names using
```{r, eval=FALSE}
vignette(package="BCWaterDischargeAnalysis")
```

You can read the vignette using:
```{r, eval=FALSE}
vignette("example-analysis", package="BCWaterDischargeAnalysis")
```


## Help on functions
Use the help function of R in the usual way.
```{r}
help(package="BCWaterDischargeAnalysis") # list of all functions in package
help(compute.Q.stat.annual)  # help on individual function
```

### Data
Create a data frame (e.g. __df__) with two varibles:

Variable | Description
---------|-------------
Date   | The data of the reading in standard R date class format. For example to convert a 4-digit year, numeric month, and numeric day of the month use __df$Date  <- as.Date{paste(df$Year,'-',df$Month,'-',df$Day,sep="")}__.
Q  | The average daily flow as a numeric value.

Other variables in the data frame will be ignored.

Missing values can be indicated by leaving out dates with missing
discharge values, or by setting the value of __Q__ to missing (NA).

The treatment of missing values in the analysis is explained in more
detail in the vignette.




### Project Status
This project is complete.

### Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/water-discharge-analysis/issues/).

### License

    Copyright 2017 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.



